Defining Transcriptome using RNA Sequencing\\ of {\em Streptococcus mutans}

### 2012-04-12 package/supplementary.Rnw
Just an update.

<<eval=FALSE,echo=FALSE,results=hide>>=
# not run when smutans R package is installed
# Use this when smutans R package is being tested.
library(DESeq)
library(goseq)
library(GenomicRanges)
library(IRanges)
library(plotrix)
library(xtable)
library(RColorBrewer)
source("smutans/R/class_and_slots.R")
source("smutans/R/core.R")
source("smutans/R/methods.R")
load("smutans/data/smutansData.tx.RData")
load("smutans/data/smutansData.txGenes.RData")
load("smutans/data/smutansData.txGrCombined.RData")
load("smutans/data/smutansData.txAnnotation.RData")
load("smutans/data/smutansData.txRNAz.RData")
load("smutans/data/smutansData.txPileup.RData")
load("smutans/data/smutans.feature.genes.RData")
load("smutans/data/smutans.genes.criteria.RData")
load("smutans/data/smutans.go.genes.RData")
load("smutans/data/smutans.cat.desc.RData")
load("smutans/data/smutans.keg.genes.RData")
load("smutans/data/smutans.keg.desc.RData")
load("smutans/data/smutansGenes.RData")
@
<<echo=FALSE,results=hide>>=
# not run without smutans R package installed
# Remove eval=FALSE when the R package is installed
library(smutans)
data(smutansData.tx)
data(smutansData.txGenes)
data(smutansData.txGrCombined)
data(smutansData.txAnnotation)
data(smutansData.txRNAz)
data(smutansData.txPileup)
data(smutans.feature.genes)
data(smutans.genes.criteria)
data(smutans.go.genes)
data(smutans.cat.desc)
data(smutans.keg.genes)
data(smutans.keg.desc)
data(smutansGenes)
data(smutansGenes2)
@

@@

%\clearpage{}
<<echo=FALSE,results=tex>>=
ir1 <- gaps(ranges(smutansData.txAnnotation))
ir2 <- ranges(smutansData.txRNAz)
mtchUnannotatedRNA <- findOverlaps(ir2,ir1,type="within")
txRNAz.unannotated <- ir2[matchMatrix(mtchUnannotatedRNA)[,1]]
t1 <- data.frame(names=names(txRNAz.unannotated),start=start(txRNAz.unannotated),end=end(txRNAz.unannotated))
x.big <- xtable( t1,
                 display=c("s","s","d","d"),
                 align=c('l','l','r','r'),
                 label='tabUnannotatedRNA',
                 caption='{\\bf List of unannotated RNAs predicted using RNAz.}'
               )
print( x.big,
       caption.placement="top",
       include.rownames=FALSE)
@

@@

matchUnannotatedRNA <- findOverlaps(grRNAz,grNoAnnotation,type="within")
txRNAz.unannotated <- grRNAz[matchMatrix(matchUnannotatedRNA)[,1]]

@@

# Find the number of genes in segments.
matchGenesTx2 <- findOverlaps(smutansData.txGenes,grTx2,type="within")
number.genes.tx <- runLength(Rle(matchMatrix(matchGenesTx2)[,"subject"]))

# Length distribution of UTR regions
# grTx2 <- grTxWGenes[-txWithConflictedGenes] 
# Lengths of 5' and 3' UTRs
mtchPlus <- findOverlaps(grGenesPlus,grTx2,type="within")
mtchMnus <- findOverlaps(grGenesMnus,grTx2,type="within")
length5UTR <- c()
length3UTR <- c()
for (i in unique(matchMatrix(mtchPlus)[,"subject"])) {
  x <- matchMatrix(mtchPlus)[matchMatrix(mtchPlus)[,"subject"]==i,"query"]
  y <- min(start(grGenesPlus[x])) - start(grTx2[i])
  z <- end(grTx2[i]) - max(end(grGenesPlus[x])) 
  length5UTR <- c(length5UTR,y)
  length3UTR <- c(length3UTR,z)
}
for (i in unique(matchMatrix(mtchMnus)[,"subject"])) {
  x <- matchMatrix(mtchMnus)[matchMatrix(mtchMnus)[,"subject"]==i,"query"]
  y <- min(start(grGenesMnus[x])) - start(grTx2[i])
  z <- end(grTx2[i]) - max(end(grGenesMnus[x])) 
  length5UTR <- c(length5UTR,z)
  length3UTR <- c(length3UTR,y)
}

# Correlation of expression levels of adjacent genes.
pileup.x <- lapply(ranges(smutansData.txGenes), function(x) seqselect(smutansData.txPileup, x))
pileup.y <- mapply(function(x,y) sum(x)/y, x=pileup.x, y=width(smutansData.txGenes))
elementMetadata(smutansData.txGenes)["score"] <- pileup.y
mtch3 <- findOverlaps(smutansData.txGenes,grTx2)
x <- rle(matchMatrix(mtch3)[,"subject"])
comp.exp.two.genes <- c()
for (i in x$values[x$lengths > 1]) {
  y <- combn(matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==i],2)
  for (j in 1:dim(y)[2]) {
    comp.exp.two.genes <- cbind(comp.exp.two.genes,as.matrix(elementMetadata(smutansData.txGenes)["score"][y[,j],]))
  }
}
# Correlation of expression levels of adjacent genes between transcripts.
comp.exp.two.genes.between.transcript <- c()
for (i in 1:length(x$values)) {
  if (i == length(x$values)) {
    break
  }
  y1 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i]]
  i2 <- i + 1
  y2 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i2]]
  l <- as.matrix(do.call(expand.grid, list(y1,y2)))
  for (j in 1:dim(l)[1]) {
    comp.exp.two.genes.between.transcript <- cbind(comp.exp.two.genes.between.transcript,as.matrix(elementMetadata(smutansData.txGenes)["score"][l[j,],]))
  }
}
@

<<figLength,fig=TRUE,echo=FALSE>>=
l <- list(expressed=width(smutansData.txGenes[matchMatrix(matchGenesTx)[,1]]), 
          notexpressed=width(smutansData.txGenes[-matchMatrix(matchGenesTx)[,1]]))
bin <- c(as.integer(seq.int(0, 3000, length.out=15)), 8500)
multhist(l,breaks=bin)
@
\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{supplementary-figLength}
\caption{{\bf Length distribution for expressed and not-expressed genes.}
Darker gray
bars represent frequency of expressed genes, and lighter gray ones that of
not-expressed genes. The last bin sums from 3000 bp to 8500 bp.
}
\label{figLength}
\end{figure}

@@

<<figDotplotGlucoseOnly,fig=TRUE,echo=FALSE,width=12,height=6>>=
#par(mar=c(4.1, 3.1, 0.1, 0.1))
par(mai=c(0, 0, 0, 0))
y <- tableGlucoseOnly[,"Functional Class"]
# z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
# z <- table(as.factor(sort(unlist(z))))
z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
xm <- tableGlucoseOnly[rep(seq(length(z)),sapply(z,length)),]
xm[,"Functional Class"] <- unlist(z)
# Use xm to plot a dotplot.
dotplot(xm[,"Functional Class"]~xm[,"fold"],
        main="",xlab="Log2 fold change",col="black",pch=88)
@
\begin{sidewaysfigure}
\centering
\includegraphics[width=0.9\textwidth]{supplementary-figDotplotGlucoseOnly}
\caption{{\bf Distribution of functional classes of differntially expressed genes
in comparing UA159 wild-type and TW1 mutant-type grown in glucose.}
}
\label{figDotplotGlucoseOnly}
\end{sidewaysfigure}

@@

<<figDotplotGalactoseOnly,fig=TRUE,echo=FALSE,width=12,height=6>>=
par(mai=c(0, 0, 0, 0))
y <- tableGalactoseOnly[,"Functional Class"]
# z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
# z <- table(as.factor(sort(unlist(z))))
z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
xm <- tableGalactoseOnly[rep(seq(length(z)),sapply(z,length)),]
xm[,"Functional Class"] <- unlist(z)
# Use xm to plot a dotplot.
dotplot(xm[,"Functional Class"]~xm[,"fold"],
        main="",xlab="Log2 fold change",col="black",pch=88)
@
\begin{sidewaysfigure}
\centering
\includegraphics[width=0.9\textwidth]{supplementary-figDotplotGalactoseOnly}
\caption{{\bf Distribution of functional classes of differntially expressed genes
in comparing UA159 wild-type and TW1 mutant-type grown in galactose.}
}
\label{figDotplotGalactoseOnly}
\end{sidewaysfigure}

@@

<<figDotplotUA159Only,fig=TRUE,echo=FALSE,width=12,height=6>>=
#par(mar=c(4.1, 3.1, 0.1, 0.1))
par(mai=c(0, 0, 0, 0))
y <- tableUA159Only[,"Functional Class"]
# z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
# z <- table(as.factor(sort(unlist(z))))
z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
xm <- tableUA159Only[rep(seq(length(z)),sapply(z,length)),]
xm[,"Functional Class"] <- unlist(z)
# Use xm to plot a dotplot.
dotplot(xm[,"Functional Class"]~xm[,"fold"],
        main="",xlab="Log2 fold change",col="black",pch=88)
@
\begin{sidewaysfigure}
\centering
\includegraphics[width=0.9\textwidth]{supplementary-figDotplotUA159Only}
\caption{{\bf Distribution of functional classes of differntially expressed genes
in comparing UA159 wild-type grown in glucose and galactose.}
}
\label{figDotplotUA159Only}
\end{sidewaysfigure}

@@

<<figDotplotTW1Only,fig=TRUE,echo=FALSE,width=12,height=6>>=
#par(mar=c(4.1, 3.1, 0.1, 0.1))
par(mai=c(0, 0, 0, 0))
y <- tableTW1Only[,"Functional Class"]
# z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
# z <- table(as.factor(sort(unlist(z))))
z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
xm <- tableTW1Only[rep(seq(length(z)),sapply(z,length)),]
xm[,"Functional Class"] <- unlist(z)
# Use xm to plot a dotplot.
dotplot(xm[,"Functional Class"]~xm[,"fold"],
        main="",xlab="Log2 fold change",col="black",pch=88)
@
\begin{sidewaysfigure}
\centering
\includegraphics[width=0.9\textwidth]{supplementary-figDotplotTW1Only}
\caption{{\bf Distribution of functional classes of differntially expressed genes
in comparing TW1 wild-type grown in glucose and galactose.}
}
\label{figDotplotTW1Only}
\end{sidewaysfigure}

>>>>

<<eval=FALSE,echo=FALSE,results=hide>>=
# not run when smutans R package is installed
# Use this when smutans R package is being tested.
library(DESeq)
library(goseq)
library(GenomicRanges)
library(IRanges)
library(plotrix)
library(xtable)
library(RColorBrewer)
source("smutans/R/class_and_slots.R")
source("smutans/R/core.R")
source("smutans/R/methods.R")
load("smutans/data/smutansData.tx.RData")
load("smutans/data/smutansData.txGenes.RData")
load("smutans/data/smutansData.txGrCombined.RData")
load("smutans/data/smutansData.txAnnotation.RData")
load("smutans/data/smutansData.txRNAz.RData")
load("smutans/data/smutansData.txPileup.RData")
load("smutans/data/smutans.feature.genes.RData")
load("smutans/data/smutans.genes.criteria.RData")
load("smutans/data/smutans.go.genes.RData")
load("smutans/data/smutans.cat.desc.RData")
load("smutans/data/smutans.keg.genes.RData")
load("smutans/data/smutans.keg.desc.RData")
load("smutans/data/smutansGenes.RData")
load("smutans/data/tw1.summary.stat.xtable.RData")
@
<<echo=FALSE,results=hide>>=
# not run without smutans R package installed
# Remove eval=FALSE when the R package is installed
library(smutans)
inDir = system.file("extdata", package="smutans", mustWork=TRUE)
smutansGenes <- 
  readSmutans (countsFile=file.path(inDir, "count-ua159-bwa.txt"),
               indexFile =file.path(inDir, "count-ua159.txt.index"),
               condition = c("UA159GLU", "UA159GAL", "TW1GLU", "TW1GAL"),
               firstFactor = c("UA159GLU", "UA159GAL"),
               firstFactorLabel = c("ua159", "tw1"),
               secondFactor = c("UA159GLU", "TW1GLU"),
               secondFactorLabel = c("glucose","galactose"),
               name="S. mutans UA159, TW1, Glucose, and Galactose")
data(smutansData.tx)
data(smutansData.txGenes)
data(smutansData.txGrCombined)
data(smutansData.txAnnotation)
data(smutansData.txRNAz)
data(smutansData.txPileup)
data(smutans.feature.genes)
data(smutans.genes.criteria)
data(smutans.go.genes)
data(smutans.cat.desc)
data(smutans.keg.genes)
data(smutans.keg.desc)
data(smutansGenes2)
data(tw1.summary.stat.xtable)
@

@@

%\clearpage{}

<<echo=FALSE,results=tex>>=
# lab:count
print( tw1.summary.stat.xtable,
       caption.placement="top",
       include.rownames=FALSE )
@

<<echo=FALSE,results=tex>>=
ir1 <- gaps(ranges(smutansData.txAnnotation))
ir2 <- ranges(smutansData.txRNAz)
mtchUnannotatedRNA <- findOverlaps(ir2,ir1,type="within")
txRNAz.unannotated <- ir2[as.matrix(mtchUnannotatedRNA)[,1]]
t1 <- data.frame(names=names(txRNAz.unannotated),start=start(txRNAz.unannotated),end=end(txRNAz.unannotated))
x.big <- xtable( t1,
                 display=c("s","s","d","d"),
                 align=c('l','l','r','r'),
                 label='tabUnannotatedRNA',
                 caption='{\\bf List of unannotated RNAs predicted using RNAz.}'
               )
print( x.big,
       caption.placement="top",
       include.rownames=FALSE)
@

@@

matchUnannotatedRNA <- findOverlaps(grRNAz,grNoAnnotation,type="within")
txRNAz.unannotated <- grRNAz[as.matrix(matchUnannotatedRNA)[,1]]

@@

# Find the number of genes in segments.
matchGenesTx2 <- findOverlaps(smutansData.txGenes,grTx2,type="within")
number.genes.tx <- runLength(Rle(as.matrix(matchGenesTx2)[,"subjectHits"]))

# Length distribution of UTR regions
# grTx2 <- grTxWGenes[-txWithConflictedGenes] 
# Lengths of 5' and 3' UTRs
mtchPlus <- findOverlaps(grGenesPlus,grTx2,type="within")
mtchMnus <- findOverlaps(grGenesMnus,grTx2,type="within")
length5UTR <- c()
length3UTR <- c()
for (i in unique(as.matrix(mtchPlus)[,"subjectHits"])) {
  x <- as.matrix(mtchPlus)[as.matrix(mtchPlus)[,"subjectHits"]==i,"queryHits"]
  y <- min(start(grGenesPlus[x])) - start(grTx2[i])
  z <- end(grTx2[i]) - max(end(grGenesPlus[x])) 
  length5UTR <- c(length5UTR,y)
  length3UTR <- c(length3UTR,z)
}
for (i in unique(as.matrix(mtchMnus)[,"subjectHits"])) {
  x <- as.matrix(mtchMnus)[as.matrix(mtchMnus)[,"subjectHits"]==i,"queryHits"]
  y <- min(start(grGenesMnus[x])) - start(grTx2[i])
  z <- end(grTx2[i]) - max(end(grGenesMnus[x])) 
  length5UTR <- c(length5UTR,z)
  length3UTR <- c(length3UTR,y)
}

# Correlation of expression levels of adjacent genes.
pileup.x <- lapply(ranges(smutansData.txGenes), function(x) seqselect(smutansData.txPileup, x))
pileup.y <- mapply(function(x,y) sum(x)/y, x=pileup.x, y=width(smutansData.txGenes))
elementMetadata(smutansData.txGenes)["score"] <- pileup.y
mtch3 <- findOverlaps(smutansData.txGenes,grTx2)
x <- rle(as.matrix(mtch3)[,"subjectHits"])
comp.exp.two.genes <- c()
for (i in x$values[x$lengths > 1]) {
  y <- combn(as.matrix(mtch3)[,"queryHits"][as.matrix(mtch3)[,"subjectHits"]==i],2)
  for (j in 1:dim(y)[2]) {
    comp.exp.two.genes <- cbind(comp.exp.two.genes,as.matrix(elementMetadata(smutansData.txGenes)["score"][y[,j],]))
  }
}
# Correlation of expression levels of adjacent genes between transcripts.
comp.exp.two.genes.between.transcript <- c()
for (i in 1:length(x$values)) {
  if (i == length(x$values)) {
    break
  }
  y1 <- as.matrix(mtch3)[,"queryHits"][as.matrix(mtch3)[,"subjectHits"]==x$values[i]]
  i2 <- i + 1
  y2 <- as.matrix(mtch3)[,"queryHits"][as.matrix(mtch3)[,"subjectHits"]==x$values[i2]]
  l <- as.matrix(do.call(expand.grid, list(y1,y2)))
  for (j in 1:dim(l)[1]) {
    comp.exp.two.genes.between.transcript <- cbind(comp.exp.two.genes.between.transcript,as.matrix(elementMetadata(smutansData.txGenes)["score"][l[j,],]))
  }
}
@

<<figLength,fig=TRUE,echo=FALSE>>=
l <- list(expressed=width(smutansData.txGenes[as.matrix(matchGenesTx)[,1]]), 
          notexpressed=width(smutansData.txGenes[-as.matrix(matchGenesTx)[,1]]))
bin <- c(as.integer(seq.int(0, 3000, length.out=15)), 8500)
multhist(l,breaks=bin)
@
\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{supplementary-figLength}
\caption{{\bf Length distribution for expressed and not-expressed genes.}
Darker gray
bars represent frequency of expressed genes, and lighter gray ones that of
not-expressed genes. The last bin sums from 3000 bp to 8500 bp.
}
\label{figLength}
\end{figure}

@@

<<figDotplotGlucoseOnly,fig=TRUE,echo=FALSE,width=12,height=6>>=
#par(mar=c(4.1, 3.1, 0.1, 0.1))
par(mai=c(0, 0, 0, 0))
y <- tableGlucoseOnly[,"Functional Class"]
# z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
# z <- table(as.factor(sort(unlist(z))))
z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
xm <- tableGlucoseOnly[rep(seq(length(z)),sapply(z,length)),]
xm[,"Functional Class"] <- unlist(z)
# Use xm to plot a dotchart.
dotchart(xm[,"fold"],labels=xm[,"Functional Class"],
        main="",xlab="Log2 fold change",col="black",pch=88)
@
\begin{sidewaysfigure}
\centering
\includegraphics[width=0.9\textwidth]{supplementary-figDotplotGlucoseOnly}
\caption{{\bf Distribution of functional classes of differntially expressed genes
in comparing UA159 wild-type and TW1 mutant-type grown in glucose.}
}
\label{figDotplotGlucoseOnly}
\end{sidewaysfigure}

@@

<<figDotplotGalactoseOnly,fig=TRUE,echo=FALSE,width=12,height=6>>=
par(mai=c(0, 0, 0, 0))
y <- tableGalactoseOnly[,"Functional Class"]
# z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
# z <- table(as.factor(sort(unlist(z))))
z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
xm <- tableGalactoseOnly[rep(seq(length(z)),sapply(z,length)),]
xm[,"Functional Class"] <- unlist(z)
# Use xm to plot a dotchart.
dotchart(xm[,"fold"],labels=xm[,"Functional Class"],
        main="",xlab="Log2 fold change",col="black",pch=88)
@
\begin{sidewaysfigure}
\centering
\includegraphics[width=0.9\textwidth]{supplementary-figDotplotGalactoseOnly}
\caption{{\bf Distribution of functional classes of differntially expressed genes
in comparing UA159 wild-type and TW1 mutant-type grown in galactose.}
}
\label{figDotplotGalactoseOnly}
\end{sidewaysfigure}

@@

<<figDotplotUA159Only,fig=TRUE,echo=FALSE,width=12,height=6>>=
#par(mar=c(4.1, 3.1, 0.1, 0.1))
par(mai=c(0, 0, 0, 0))
y <- tableUA159Only[,"Functional Class"]
# z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
# z <- table(as.factor(sort(unlist(z))))
z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
xm <- tableUA159Only[rep(seq(length(z)),sapply(z,length)),]
xm[,"Functional Class"] <- unlist(z)
# Use xm to plot a dotchart.
dotchart(xm[,"fold"],labels=xm[,"Functional Class"],
        main="",xlab="Log2 fold change",col="black",pch=88)
@
\begin{sidewaysfigure}
\centering
\includegraphics[width=0.9\textwidth]{supplementary-figDotplotUA159Only}
\caption{{\bf Distribution of functional classes of differntially expressed genes
in comparing UA159 wild-type grown in glucose and galactose.}
}
\label{figDotplotUA159Only}
\end{sidewaysfigure}

@@

<<figDotplotTW1Only,fig=TRUE,echo=FALSE,width=12,height=6>>=
#par(mar=c(4.1, 3.1, 0.1, 0.1))
par(mai=c(0, 0, 0, 0))
y <- tableTW1Only[,"Functional Class"]
# z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
# z <- table(as.factor(sort(unlist(z))))
z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
xm <- tableTW1Only[rep(seq(length(z)),sapply(z,length)),]
xm[,"Functional Class"] <- unlist(z)
# Use xm to plot a dotchart.
dotchart(xm[,"fold"],labels=xm[,"Functional Class"],
        main="",xlab="Log2 fold change",col="black",pch=88)
@
\begin{sidewaysfigure}
\centering
\includegraphics[width=0.9\textwidth]{supplementary-figDotplotTW1Only}
\caption{{\bf Distribution of functional classes of differntially expressed genes
in comparing TW1 wild-type grown in glucose and galactose.}
}
\label{figDotplotTW1Only}
\end{sidewaysfigure}
